Interpretability and Fairness in Machine Learning: A Formal Methods Approach

### 2023-02-10 chapters/fairness/fif/conclusion.tex
fix bib

In this paper, we propose the Fairness Influence Function (FIF) to measure the effect of input features on the bias of classifiers on a given dataset. Our approach combines global sensitivity analysis and group-based fairness metrics in machine learning. Thereby, it is natural in our approach to formulate FIF of intersectional features, which together with individual FIFs explains bias with higher granularity. We theoretically analyze the properties of FIFs and provide an algorithm, {\fairXplainer}, for estimating FIFs using global variance decomposition and local regression. In experiments, {\fairXplainer} estimates individual and intersectional FIFs in real-world datasets and classifiers, approximates bias using FIFs with less estimation error than earlier methods, demonstrates a high correlation between FIFs and fairness interventions, and analyzes the impact of fairness enhancing and attack algorithms on FIFs. The results instantiate {\fairXplainer} as a global, granular, and more accurate explanation method to understand the sources of bias. Additionally, the resonance between the rules extracted by CORELS~\cite{rudin19} and the most influential features detected by {\fairXplainer} indicates that FIFs can be exploited in future to create an explainable proxy of a biased/unbiased classifier. Also, we aim to extend {\fairXplainer} to compute FIFs for complex data, such as image and text, and design algorithms leveraging FIFs to yield unbiased decisions.
%

>>>>

In this paper, we propose the Fairness Influence Function (FIF) to measure the effect of input features on the bias of classifiers on a given dataset. Our approach combines global sensitivity analysis and group-based fairness metrics in machine learning. Thereby, it is natural in our approach to formulate FIF of intersectional features, which together with individual FIFs explains bias with higher granularity. We theoretically analyze the properties of FIFs and provide an algorithm, {\fairXplainer}, for estimating FIFs using global variance decomposition and local regression. In experiments, {\fairXplainer} estimates individual and intersectional FIFs in real-world datasets and classifiers, approximates bias using FIFs with less estimation error than earlier methods, demonstrates a high correlation between FIFs and fairness interventions, and analyzes the impact of fairness enhancing and attack algorithms on FIFs. The results instantiate {\fairXplainer} as a global, granular, and more accurate explanation method to understand the sources of bias. Additionally, the resonance between the rules extracted by CORELS~\cite{rudin19stop} and the most influential features detected by {\fairXplainer} indicates that FIFs can be exploited in future to create an explainable proxy of a biased/unbiased classifier. Also, we aim to extend {\fairXplainer} to compute FIFs for complex data, such as image and text, and design algorithms leveraging FIFs to yield unbiased decisions.
%

### 2023-02-10 chapters/fairness/fif/experiments.tex
fix bib

\subsection{Explainability and Applicability of FIFs}
\paragraph{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs in Figure~\ref{fairness_fairXplainer_fig:fif_illustration_compas}. We consider COMPAS dataset with race $ \in $ \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years. Since the classifier only optimizes training error, it demonstrates statistical parity of $ 0.174 $, i.e. it suggests that a non-Caucasian  has $ 0.174 $ higher probability of re-offending crimes than a Caucasian. Next, we investigate the sources of bias and present individual FIFs in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, and both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs}. In both figures, we present influential features and their FIFs in the descending order of absolute values. According to {\fairXplainer}, the priors count (FIF $ = 0.1 $) dominates in increasing statistical parity\textemdash between Caucasian and non-Caucasian, their prior count demonstrates the maximum difference in the scaled variance of positive prediction. Other non-sensitive features have almost zero FIFs. However, in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, higher-order FIFs ($ \lambda > 1 $) increases statistical parity by $ 0.031 $, denoting that the data is correlated and presenting only individual FIFs is not sufficient for understanding the sources of bias. For example, while both sex and age individually demonstrate almost zero influence on bias (Figure~\ref{fairness_fairXplainer_fig:individual_fifs}), their combined effect and intersectional effects with c-charge degree, priors count, and juvenile miscellaneous count contribute highly on statistical parity. In contrast to {\fairXplainer}, SHAP only estimates individual FIFs  (Figure~\ref{fairness_fairXplainer_fig:fif_shap}) and approximates statistical parity with higher error than {\fairXplainer}.  Interestingly, {\fairXplainer} yields FIF estimates of the classifier trained on COMPAS dataset that significantly matches with the rule-based classifier extracted from the COMPAS dataset using Certifiably Optimal Rule Lists (CORELS) algorithm~\cite[Figure 3]{rudin19}. From Figure 3 in~\cite{rudin19}, we observe that prior count is the only feature used individually to predict arrests, while age is paired with sex and prior counts respectively. While considering second-order intersectionality, {\fairXplainer} ($ \lambda = 2 $) yields priors count, (sex, age), and (age, priors count) as \emph{three of the top four features} that explains the observation of~\cite{rudin19} better than SHAP and {\fairXplainer} ($ \lambda = 1 $) considering only individual features. \emph{Therefore, {\fairXplainer} demonstrates a clearer understanding on the sources of bias of a classifier by simultaneously quantifying on intersectional influences and individual influences }.  %\todo{For a linear classifier with independent features, see if coefficient and FIFs are related.}

>>>>

\subsection{Explainability and Applicability of FIFs}
\paragraph{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs in Figure~\ref{fairness_fairXplainer_fig:fif_illustration_compas}. We consider COMPAS dataset with race $ \in $ \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years. Since the classifier only optimizes training error, it demonstrates statistical parity of $ 0.174 $, i.e. it suggests that a non-Caucasian  has $ 0.174 $ higher probability of re-offending crimes than a Caucasian. Next, we investigate the sources of bias and present individual FIFs in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, and both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs}. In both figures, we present influential features and their FIFs in the descending order of absolute values. According to {\fairXplainer}, the priors count (FIF $ = 0.1 $) dominates in increasing statistical parity\textemdash between Caucasian and non-Caucasian, their prior count demonstrates the maximum difference in the scaled variance of positive prediction. Other non-sensitive features have almost zero FIFs. However, in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, higher-order FIFs ($ \lambda > 1 $) increases statistical parity by $ 0.031 $, denoting that the data is correlated and presenting only individual FIFs is not sufficient for understanding the sources of bias. For example, while both sex and age individually demonstrate almost zero influence on bias (Figure~\ref{fairness_fairXplainer_fig:individual_fifs}), their combined effect and intersectional effects with c-charge degree, priors count, and juvenile miscellaneous count contribute highly on statistical parity. In contrast to {\fairXplainer}, SHAP only estimates individual FIFs  (Figure~\ref{fairness_fairXplainer_fig:fif_shap}) and approximates statistical parity with higher error than {\fairXplainer}.  Interestingly, {\fairXplainer} yields FIF estimates of the classifier trained on COMPAS dataset that significantly matches with the rule-based classifier extracted from the COMPAS dataset using Certifiably Optimal Rule Lists (CORELS) algorithm~\cite[Figure 3]{rudin19stop}. From Figure 3 in~\cite{rudin19stop}, we observe that prior count is the only feature used individually to predict arrests, while age is paired with sex and prior counts respectively. While considering second-order intersectionality, {\fairXplainer} ($ \lambda = 2 $) yields priors count, (sex, age), and (age, priors count) as \emph{three of the top four features} that explains the observation of~\cite{rudin19stop} better than SHAP and {\fairXplainer} ($ \lambda = 1 $) considering only individual features. \emph{Therefore, {\fairXplainer} demonstrates a clearer understanding on the sources of bias of a classifier by simultaneously quantifying on intersectional influences and individual influences }.  %\todo{For a linear classifier with independent features, see if coefficient and FIFs are related.}

### 2023-02-10 doctoral_consortium/main.tex
fix bib

\subsection*{Probabilistic Fairness Verification} The problem in probabilistic fairness verification is to verify the bias of a classifier given the distribution of input features. The early works on fairness verification focused on measuring fairness metrics of a classifier for a given dataset~\cite{aif360-oct-2018}. Naturally, such techniques were limited in enhancing confidence of users for wide deployment. Consequently, recent verifiers seek to achieve verification beyond  finite dataset and in turn focus on the  probability distribution of features~\cite{albarghouthi2017fairsquare, bastani2019probabilistic}.  More specifically, the input to the verifier is a classifier and  the probability distribution of features, and the output is an estimate of fairness metrics that the classifier obtains given the distribution.

@@

Linear classifiers have attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. Existing fairness verifier suffers from two-fold limitations for verifying linear classifiers: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations. Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification. {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming. To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network to represent the conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG). Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms. We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing fairness influences of a subset of features on shifting the incurred bias of the classifiers from the original bias.
%, which we study in detail in the following contribution of the thesis.

>>>>

\subsection*{Probabilistic Fairness Verification} The problem in probabilistic fairness verification is to verify the bias of a classifier given the distribution of input features. The early works on fairness verification focused on measuring fairness metrics of a classifier for a given dataset~\cite{aif360-oct-2018}. Naturally, such techniques were limited in enhancing confidence of users for wide deployment. Consequently, recent verifiers seek to achieve verification beyond  finite dataset and in turn focus on the  probability distribution of features~\cite{albarghouthi2017fairsquare,bastani2019probabilistic}.  More specifically, the input to the verifier is a classifier and  the probability distribution of features, and the output is an estimate of fairness metrics that the classifier obtains given the distribution.

@@

Linear classifiers have attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy,john2020verifying}. Existing fairness verifier suffers from two-fold limitations for verifying linear classifiers: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations. Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification. {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming. To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network to represent the conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG). Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms. We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing fairness influences of a subset of features on shifting the incurred bias of the classifiers from the original bias.
%, which we study in detail in the following contribution of the thesis.
